\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-01-21 analysis/paper/paper.tex

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R>                        levels = c(0, 1), 
R>                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R>                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R>                                     time = seq(0,15,0.1), 
R>                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

@@

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"), # Both diseases
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R>                           const(Phase) + const(Source) + const(Age), 
R>                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

\caption{\label{fig:support-poptime} Population-time plot for tje SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R>                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R>                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R>                   txtime = time_length(accept.dt %--% tx.date, 
R>                                        unit = "days"),
R>                   exposure = case_when(
R>                     is.na(txtime) ~ 0L,
R>                     txtime > futime ~ 0L,
R>                     txtime <= futime ~ 1L
R>                   ))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R>                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R>                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R>                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R>                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

@@

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R>   "Model2" = AIC(fit2),
R>   "Model3" = AIC(fit3),
R>   "Model4" = AIC(fit4))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R>                            futime = seq(0, 1000,
R>                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R>                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R>                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R>     geom_line() +
R>     theme_minimal() +
R>     theme(legend.position = 'top') +
R>     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
R>                         time = seq(0, 1000, length.out = 100))
R> 
R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
R> 
R> # Rearrange the data
R> absrisk <- gather(as.data.frame(absrisk),
R>                   "Status", "Risk", -Time)
R>  
R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
R>   geom_line() +
R>   theme_minimal() +
R>   theme(legend.position = 'top') +
R>   expand_limits(y = 0.5) +
R>   xlab('Follow-up time') + ylab('Cum. Incidence')
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> one_yr_haz <- data.frame(futime = seq(0, 365,
R>                                       length.out = 100))
R> one_yr_haz <- mutate(one_yr_haz,
R>                      offset = 0,
R>                      exposure = if_else(futime < 30, 0, 1))
R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
R>                                 type = "link"))
R> ggplot(one_yr_haz, aes(futime, hazard)) +
R>   geom_line() +
R>   theme_minimal() +
R>   theme(legend.position = 'top') +
R>   ylab('Hazard') + xlab('Follow-up time') +
R>   geom_vline(xintercept = 30, linetype = 'dashed')
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
R>              time = 365)
R> 
R> # Use the trapezoidal rule to estimate the cumulative hazard
R> min_hazard <- min(one_yr_haz$hazard)
R> max_hazard <- max(one_yr_haz$hazard)
R> increment <- 365/99
R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
R> 1 - exp(-cumhaz_est)
\end{CodeInput}
\end{CodeChunk}

@@

In this article, we presented the \proglang{R} package \pkg{casebase}
that provides functions to fit smooth parametric hazards and estimate
cumulative incidence functions using case-base sampling. We outlined the
theoretical underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through three case studies.

\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
censoring, whereas \pkg{casebase} currently cannot. Interval censoring
is a simple feature to implement and is a potential improvement to add
to \pkg{casebase}.

>>>>

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0, 1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}

@@

\begin{CodeChunk}
\begin{table}[H]
\centering
\begin{tabular}{lrlr}
\toprule
Model & HR & CI & Width\\
\midrule
Cox & 0.80 & (0.67, 0.95) & 0.28\\
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
Weibull & 0.79 & (0.67, 0.94) & 0.28\\
Splines & 0.81 & (0.68, 0.96) & 0.28\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5790  3      262 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R+   "Weib." = AIC(casebase_time),
R+   "Splines" = AIC(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>    Exp.   Weib. Splines 
#>    6057    5817    5800
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics{../figures/compPop-1}

\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.}\label{fig:compPop}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R+                                      levels = levels(bmtcrr[,"Sex"])),
R+                       "D" = c("ALL", "AML"), # Both diseases
R+                       "Phase" = factor(c("Relapse", "Relapse"), 
R+                                        levels = levels(bmtcrr[,"Phase"])),
R+                       "Age" = c(35, 35),
R+                       "Source" = factor(c("PB", "PB"), 
R+                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R+                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R+                           const(Phase) + const(Source) + const(Age), 
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R+                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}

@@

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R+                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R+                   txtime = time_length(accept.dt %--% tx.date, 
R+                                        unit = "days"),
R+                   exposure = case_when(
R+                     is.na(txtime) ~ 0L,
R+                     txtime > futime ~ 0L,
R+                     txtime <= futime ~ 1L
R+                   ))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R+                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R+                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

@@

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R+   "Model2" = AIC(fit2),
R+   "Model3" = AIC(fit3),
R+   "Model4" = AIC(fit4))
\end{CodeInput}

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    795    760    729    724
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R+                            futime = seq(0, 1000,
R+                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R+                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R+                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+     geom_line() +
R+     theme_minimal() +
R+     theme(legend.position = 'top') +
R+     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}

@@

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
R+                         time = seq(0, 1000, length.out = 100))
R> 
R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
R> 
R> # Rearrange the data
R> absrisk <- gather(as.data.frame(absrisk),
R+                   "Status", "Risk", -Time)
R>  
R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
R+   geom_line() +
R+   theme_minimal() +
R+   theme(legend.position = 'top') +
R+   expand_limits(y = 0.5) +
R+   xlab('Follow-up time') + ylab('Cum. Incidence')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> one_yr_haz <- data.frame(futime = seq(0, 365,
R+                                       length.out = 100))
R> one_yr_haz <- mutate(one_yr_haz,
R+                      offset = 0,
R+                      exposure = if_else(futime < 30, 0, 1))
R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
R+                                 type = "link"))
R> ggplot(one_yr_haz, aes(futime, hazard)) +
R+   geom_line() +
R+   theme_minimal() +
R+   theme(legend.position = 'top') +
R+   ylab('Hazard') + xlab('Follow-up time') +
R+   geom_vline(xintercept = 30, linetype = 'dashed')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}

@@

\begin{CodeInput}
R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
R+              time = 365)
\end{CodeInput}

\begin{CodeOutput}
#>         
#> 365 0.27
\end{CodeOutput}

\begin{CodeInput}
R> 
R> # Use the trapezoidal rule to estimate the cumulative hazard
R> cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
R> 1 - exp(-cumhaz_est)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 0.19
\end{CodeOutput}
\end{CodeChunk}

@@

In this article, we presented the \proglang{R} package \pkg{casebase}
that provides functions to fit smooth parametric hazards and estimate
cumulative incidence functions using case-base sampling. We outlined the
theoretical underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we explored this flexibility along two fronts. On
the one hand, we showed how splines could be used as part of the linear
predictor to model the effect of time on the hazard. This strategy
yielded estimates of the survival function that were qualitatively
similar to semiparametric estimates derived from Cox regression;
however, case-base sampling led to estimates that \emph{varying smoothly
in time}. On the other hand, we also displayed the flexibility of
case-base sampling by showing how it could be combined with penalized
logistic regression to perform variable selection. Similarly, our
package \pkg{casebase} provides the user with the functionality of
Generalized Additive Models. Moreover, case-base sampling can be
combined with quasi-likelihood estimation to fit survival models that
potentially account for the presence of over-dispersion. All of the
examples illustrate how the case-base sampling framework in general, and
the package \pkg{casebase} in particular, allows the user to fit a broad
family of survival functions.

As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
case-base sampling is comprised of three steps: 1) sampling a case
series and a base series from the study; 2) fit the log-hazard as a
linear function of predictors (including time); and 3) use the fitted
hazard to estimate the cumulative incidence curve. Accordingly, our
package provides functions for each step. Moreover, the simple interface
of the \code{fittingSmoothHazard} function resembles the \code{glm}
interface, which should look familiar to new users. Our modular approach
also provides a convenient way to extend our package for new sampling or
fitting strategies.

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
the left-hand side that captures both the event type and time. Both
approaches to modeling yield user-friendly code. In terms of output,
both approaches differ significantly. Case-base sampling produces smooth
hazard and smooth cumulative incidence curves, whereas Cox regression
and Fine-Gray models produce step-wise cumulative incidence functions
and never explicitly model the hazard function. Qualitatively, we showed
that by using splines in the linear predictor, all three models yielded
similar curves. However, the smooth nature of the output of
\pkg{casebase} improves interpretation for consumers of these
predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
comparison between the Cox model and case-base sampling.
